---
title: "t-tests done the `t.test` way and the manual way"
---

```{r, echo=FALSE}
options(scipen=99999)
```

Here's a simple t-test, where we assume equal variances.

```{r}
x <- c(179.6, 171.6, 174.1, 175.1, 174.9, 171.3, 177.2, 171.6, 174.5)
y <- c(162.8, 165.5, 161.8, 168.9, 166.8, 165.5, 167.2)

t_test_result <- t.test(x, y, var.equal = TRUE)

t_test_result
```

From this, we can see the t statistics, the p-value etc.
We can also pull them out from the `t_test_result` object as follows:
```{r}
t_test_result$statistic # t-statistic
t_test_result$p.value   # p-value
t_test_result$stderr    # standard error
```

Now, let us check all of this by doing it out the long way.
```{r}
x_bar <- mean(x) # Calculate means
y_bar <- mean(y)

n_x <- length(x) # Calculate sample sizes
n_y <- length(y)

s2_x <- var(x) # Calculate variances
s2_y <- var(y)

# Calculate standard error 
Sp <- sqrt( ((n_x - 1) * s2_x + (n_y - 1) * s2_y)/(n_x + n_y - 2))
se <- Sp * sqrt(1/n_x + 1/n_y)
```

So, so far, we have the differences of the means as follows:
```{r}
x_bar - y_bar
```
And the standard error is 
```{r}
se
```
And so the t-statistic is
```{r}
# t-statistic 
t_stat <- (x_bar - y_bar)/se
t_stat
```

So, so far, so good, and all in line with the `t.test` result above.

Now, we use the cumulative distribution function `pt` to find the probability of a values as or more extreme than `r t_stat` in a t-distribution with `r n_x + n_y - 2` degrees of freedom.

```{r}
# Probability of value lower than minus t_stat
pt(-t_stat, n_x + n_y - 2)

# Probability of value greater than t_stat
pt(t_stat, n_x + n_y - 2, lower.tail = F) 
```
```{r echo=T}
# The p-value is defined as the probability
# of a result as or more extreme than the
# observed statistic ...
p_value <- pt(-t_stat, n_x + n_y - 2) + 
           pt(t_stat, n_x + n_y -2 , lower.tail = F)

p_value
```

